---
title: "Trace cross-product calculations"
author: "John Palowitch, Miheer Dewaskar"
date: "May 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Setting parameters for the simulation, and generating data:

```{r sim-params}
n <- 500
nY <- 2
rho <- 0.5
m <- 100
Beta <- 0
s2 <- 1
Sig <- diag(1 - rho, n) + matrix(rep(rho, n^2), ncol = n)

# Data generation
source("mvrnormR.R")
X <- mvrnormR(m, rep(0, n), Sig)
Y <- Beta * rowSums(X) + matrix(rnorm(m * nY, sd = sqrt(s2)), ncol = nY)
```

Doing general calculations that can be used. We should add to these as more pre-computations become needed.

```{r general-calcs}

# General Calcs
X <- scale(X)
Y <- scale(Y)

X2 <- X^2
X3 <- X^3
X4 <- X^4
Y2 <- Y^2
Y3 <- Y^3
Y4 <- Y^4


tX <- t(X)
tY <- t(Y)
tX2 <- t(X2)
tX3 <- t(X3)
XXt <- tcrossprod(X)
XXt2 <- XXt^2

Y4ColSums <- colSums(Y4)
X4ColSums <- colSums(X4)
X1RowSums <- rowSums(X)
X2RowSums <- rowSums(X2)

# Cor calcs
xyCors <- crossprod(X, Y) / (m - 1)
xyCors2 <- xyCors^2
rX <- crossprod(tX, xyCors)
r2X <- crossprod(tX, xyCors2)
r2X2 <- crossprod(tX2, xyCors2)
rX3 <- crossprod(tX3, xyCors)

```


The following is a sandbox to calculate various cross-terms and check them. Here is the term we need to square:

\begin{align*} 
S_{jk} & = r_{uujk} + r_{uj}r_{uk}(r_{uuuu} + r_{uujj} + r_{uukk} + r_{jjkk}) / 4\\
& -\;r_{uj}(r_{uuuk} + r_{ujjk}) / 2 - r_{uk}(r_{uuuj} + r_{ukkj}) / 2
\end{align*}

I have written a function to calculate 9 |B| x |B| matrices, containing the components of the above sum, returned in a list. The list contains the elements A ... C4, which correspond to $S_jk$:
\begin{align*}
S &= A + (B1 + B2 + B3 + B4) / 4 \\
&- (C1 + C2) / 2- (C3 + C4) / 2
\end{align*}

The usage of the function is:
```{r mlist}
source("tracecalcs.R")
rm(trace_uni_fast, trace_uni)
mlist <- trace_uni_mlist(Y[ , 1], X)
mlists <- lapply(1:nY, function (i) trace_uni_mlist(Y[ , i], X))
```
...it does not take the full Y matrix. This function is useful to check the accuracy of cross-term calculation. As a first example, let's try to compute the sum over $j$ and $k$ of $A^2$:

```{r Asquared}
AAs <- lapply(1:nY, function (i) {
  crossprod(Y2[ , i], crossprod(XXt2, Y2[ , i]))
})
unlist(AAs)
unlist(lapply(mlists, function (L) sum(L$A * L$A)))
```

Unfortunately, not all cross terms are as nice. The cross-terms can be divided into 3 categories:

1. Symmetric: $AA = r_{uujk}r_{uujk}$
2. Non-symmetric, half-pairwise: $AB2 = r_{uujk}r_{uj}r_{uk}r_{uujj}$
3. Non-symmetric, full-pairwise: $AC2 = r_{uujk}r_{uj}r_{ujjk}$

Cross-terms of category (2) can be obtained by pre-computing the non-pairwise term. Cross-terms of category (3) are more difficult to deal with. I will attempt AC2:

```{r AC2}
AC2s <- lapply(1:nY, function (i) {
  AC2mat <- crossprod(xyCors[ , i] * tX, tX2)
  AC2 <- crossprod(Y[ , i], crossprod(AC2mat * XXt, Y2[ , i]))
  return(AC2)
})
unlist(AC2s)
unlist(lapply(mlists, function (L) sum(L$A * L$C2)))
```

I will list all the cross-terms below. If there's no code assume there is no solution. The goal is to get a vectorized version of each term, either with `lapply` or fancy matrix maniuplation. So, all results should return a vector of length `nY` (or a one by `nY` matrix). We can check them by something similar to the last line above.


```{r first-calcs}

#AA
AAs <- lapply(1:nY, function (i) {
  crossprod(Y2[ , i], crossprod(XXt2, Y2[ , i]))
})
unlist(AAs)
unlist(lapply(mlists, function (L) sum(L$A * L$A)))



#AB1
AB1s <- lapply(1:nY, function (i) {
  Y4ColSums[i] * crossprod(Y2[ , i], rX[ , i]^2)
})
unlist(AB1s)
unlist(lapply(mlists, function (L) sum(L$A * L$B1)))



#AB2




#AB3
#AB3 <- AB2



```

Miheer, I will leave it to you to begin working on vectorizing the rest. Many of these are written for just the first $Y$ vector. Feel free to deviate from the code if you see a more efficient way. Anyway, I cannot fully guarantee their accuracy as the format of the pre-computations may have shifted slightly when I added another $Y$ vector. Be sure to add a check with `mlists` after each term.

```{r second-calcs}

#AB4




#AC1
AC1vec1 <- crossprod(Y2 * rX, X)
AC1vec2 <- crossprod(X, Y3)
AC1 <- -AC1vec1 %*% AC1vec2

#AC2
AC2 <- AC1



#AC3




#AC4
#AC4 <- AC3



#B1^2
B1B1 <- Y4ColSums^2 * sum(xyCors2)^2



#B1B2
B1B2 <- Y4ColSums * crossprod(Y2, r2X2) * sum(xyCors2)



#B1B3
B1B3 <- B1B2



#B1B4
B1B4 <- Y4ColSums * crossprod(r2X2)



#B1C1
B1C1 <- -Y4ColSums * sum(xyCors2) * crossprod(Y3, rX)



#B1C2
B1C2 <- B1C1



#B1C3
B1C3 <- -Y4ColSums * crossprod(Y, rX * r2X2)



#B1C4
B1C4 <- B1C3



#B2^2



#B2B3
B2B3 <- crossprod(Y2, r2X2)^2



#B2B4




#B2C1




#B2C2




#B2C3




#B2C4




#B3^2
#B3B3 <- B2B2



#B3B4
#B3B4 <- B2B4



#B3C1
#B3C1 <- B2C2



#B3C2
#B3C2 <- B2C1



#B3C3
#B3C3 <- B2C4



#B3C4
#B3C4 <- B2C3



#B4^2




#B4C1




#B4C2
#B4C2 <- B4C1



#B4C3




#B4C4
#B4C4 <- B4C3



#C1^2
C1C1 <- sum(xyCors2) * crossprod(crossprod(X, Y3)) / 4



#C1C2
C1C2 <- crossprod(Y3, rX)^2 / 4



#C1C3
C1C3 <- crossprod(crossprod(X, Y3), crossprod(X, Y * r2X2)) / 4



#C1C4




#C2^2
#C2C2 <- C1C1



#C2C3
#C2C3 <- C1C4



#C2C4
#C2C4 <- C1C3



#C3^2




#C3C4




#C4^2
#C4C4 <- C3C3
  
```