---
title: "Trace cross-product calculations"
author: "John Palowitch, Miheer Dewaskar"
date: "May 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Setting parameters for the simulation, and generating data:

```{r sim-params}
n <- 500
nY <- 2
rho <- 0.5
m <- 100
Beta <- 0
s2 <- 1
Sig <- diag(1 - rho, n) + matrix(rep(rho, n^2), ncol = n)

# Data generation
set.seed(12345)
source("mvrnormR.R")
X <- mvrnormR(m, rep(0, n), Sig)
Y <- Beta * rowSums(X) + matrix(rnorm(m * nY, sd = sqrt(s2)), ncol = nY)
```

Doing general calculations that can be used. We should add to these as more pre-computations become needed.

```{r general-calcs}

# General Calcs
X <- scale(X)
Y <- scale(Y)

X2 <- X^2
X3 <- X^3
X4 <- X^4
Y2 <- Y^2
Y3 <- Y^3
Y4 <- Y^4


tX <- t(X)
tY <- t(Y)
tY2 <- t(Y2)
tX2 <- t(X2)
tX3 <- t(X3)


Y4ColSums <- colSums(Y4)
X4ColSums <- colSums(X4)
X1RowSums <- rowSums(X)
X2RowSums <- rowSums(X2)


# Cor calcs
xyCors <- crossprod(X, Y) / (m - 1)
xyCors2 <- xyCors^2
xyCors2ColSums <- colSums(xyCors2)
rX <- crossprod(tX, xyCors)
r2X <- crossprod(tX, xyCors2)
r2X2 <- X2 %*% xyCors2 
rX3 <- crossprod(tX3, xyCors)
rX2 <- crossprod(tX2, xyCors)

r <- crossprod(Y,X) / (m - 1) #r <- t(xyCor)[A,]
r2 <- crossprod(Y2, X2) / (m - 1) # similar to above.
r_squared <- r^2
r3 <- crossprod(Y3,X) / (m - 1)
r3X <- tcrossprod(X, r3)
rr3X2 <- tcrossprod(X2, r*r3)
```

These calculations have to be done for every set $A$:
```{r localcalcs}
XXt <- tcrossprod(X)
XXt2 <- XXt^2
XX2t <- X %*% tX2
```

The following is a sandbox to calculate various cross-terms and check them. Here is the term we need to square:

\begin{align*} 
S_{jk} & = r_{uujk} + r_{uj}r_{uk}(r_{uuuu} + r_{uujj} + r_{uukk} + r_{jjkk}) / 4\\
& -\;r_{uj}(r_{uuuk} + r_{ujjk}) / 2 - r_{uk}(r_{uuuj} + r_{ukkj}) / 2
\end{align*}

I have written a function to calculate 9 |B| x |B| matrices, containing the components of the above sum, returned in a list. The list contains the elements A ... C4, which correspond to $S_{jk}$:
\begin{align*}
S &= A + (B1 + B2 + B3 + B4) / 4 \\
&- (C1 + C2) / 2- (C3 + C4) / 2
\end{align*}

The usage of the function is:
```{r mlist}
source("tracecalcs.R")
rm(trace_uni)
mlist <- trace_uni_mlist(Y[ , 1], X)
mlists <- lapply(1:nY, function (i) trace_uni_mlist(Y[ , i], X))

#The string terms
sb <- "r_{uj}r_{uk}"
sTerms <- list(A = "r_{uujk}",
              B1 = paste(sb, "r_{uuuu}"),
              B2 = paste(sb, "r_{uujj}"),
              B3 = paste(sb, "r_{uukk}"),
              B4 = paste(sb, "r_{jjkk}"),
              C1 = "r_{uj}r_{uuuk}",
              C2 = "r_{uj}r_{ujjk}",
              C3 = "r_{uk}r_{uuuj}",
              C4 = "r_{uk}r_{ukkj}")

result <- function(t1,t2) {
  sapply(mlists, function (L) sum(L[[t1]] * L[[t2]]))
}

term <- function(t1,t2) {
  sprintf("\\sum_{j, k} %s %s", sTerms[[t1]], sTerms[[t2]])
}

check <- function(t1, t2, val){
  all.equal(val,result(t1,t2))
}

```
...it does not take the full Y matrix. This function is useful to check the accuracy of cross-term calculation. As a first example, let's try to compute the sum over $j$ and $k$ of $A^2$:

```{r Asquared}
AAs <- sapply(1:nY, function (i) {
  crossprod(Y2[ , i], crossprod(XXt2, Y2[ , i]))
})
check("A","A", AAs)
```

**Should we normalize the terms by $(m-1)^2$?**

Unfortunately, not all cross terms are as nice. The cross-terms can be divided into 3 categories:

1. Symmetric: $AA = r_{uujk}r_{uujk}$
2. Non-symmetric, half-pairwise: $AB2 = r_{uujk}r_{uj}r_{uk}r_{uujj}$
3. Non-symmetric, full-pairwise: $AC2 = r_{uujk}r_{uj}r_{ujjk}$

Cross-terms of category (2) can be obtained by pre-computing the non-pairwise term. Cross-terms of category (3) are more difficult to deal with. I will attempt AC2:

```{r AC2-first}
AC2s <- sapply(1:nY, function (i) {
  AC2mat <- crossprod(r[i,] * tX, tX2)
  AC2 <- crossprod(Y[ , i], crossprod(AC2mat * XXt, Y2[ , i]))
  return(AC2)
})
check("A","C2", AC2s)
```

I will list all the cross-terms below. If there's no code assume there is no solution. The goal is to get a vectorized version of each term, either with `lapply` or fancy matrix maniuplation. So, all results should return a vector of length `nY` (or a one by `nY` matrix). We can check them by something similar to the last line above.

$`r term("A","A")`$
```{r AA}
#AA
AAs <- sapply(1:nY, function (i) {
  crossprod(Y2[ , i], crossprod(XXt2, Y2[ , i]))
})
check("A","A", AAs)
```
$`r term("A","B1")`$
```{r AB1}
#AB1
AB1s <- sapply(1:nY, function (i) {
  Y4ColSums[i] * crossprod(Y2[ , i], rX[ , i]^2)
})
check("A","B1", AB1s)
```

$`r term("A","B2")`$
```{r AB2}
#AB2
#
Z <- tY2 * tcrossprod(r2 * r, X)
AB2s <- sapply(1:nY, function(i){
  crossprod(Y[,i], XXt) %*% Z[i,]
})
check("A","B2",AB2s)
```

$`r term("A","B3")`$
```{r AB3}
#AB3
AB3s <- AB2s
check("A","B3", AB3s)
```

Miheer, I will leave it to you to begin working on vectorizing the rest. Many of these are written for just the first $Y$ vector. Feel free to deviate from the code if you see a more efficient way. Anyway, I cannot fully guarantee their accuracy as the format of the pre-computations may have shifted slightly when I added another $Y$ vector. Be sure to add a check with `mlists` after each term.


$`r term("A","B4")`$
```{r AB4}
#AB4
AB4s <- sapply(1:nY, function(u){
  colSums((X2 %*% (tX * xyCors[,u]))^2) %*% Y2[,u]
})
check("A","B4", AB4s)
```

$`r term("A","C1")`$
```{r AC1}

#AC1
AC1vec1 <- crossprod(Y2 * rX, X)
AC1vec2 <- crossprod(X, Y3)
AC1s <- sapply(1:nY, function(u) {
  AC1vec1[u,] %*% AC1vec2[,u]
})
check("A","C1", AC1s)
```

$`r term("A","C2")`$
```{r AC2}
#AC2

#TODO: Reuse this common computation from AB4
AC2s <- sapply(1:nY, function(u){
  crossprod(Y[,u], (X2 %*% (tX * xyCors[,u])) * XXt) %*% Y2[,u]
})
check("A","C2", AC2s)
```

$`r term("A","C3")`$
```{r AC3}
#AC3
AC3s <- AC1s
check("A","C3", AC3s)
```

$`r term("A","C4")`$
```{r AC4}
#AC4
AC4s <- AC2s
check("A","C4", AC4s)
```

$`r term("B1","B1")`$
```{r B1B1}
#B1^2
B1B1s <- Y4ColSums^2 * xyCors2ColSums^2
check("B1","B1", B1B1s)
```

$`r term("B1","B2")`$
```{r B1B2}
#B1B2
B1B2s <- sapply(1:nY, function(u){
  Y4ColSums[u] * (r2[u,] %*% xyCors2[,u]) * xyCors2ColSums[u]
}) * (m-1) #Because mlist forgets the normalization.
check("B1","B2", B1B2s)
```

$`r term("B1","B3")`$
```{r B1B3}
#B1B3
B1B3s <- B1B2s
check("B1","B3", B1B3s)
```

$`r term("B1","B4")`$
```{r B1B4}
#B1B4
B1B4s <- Y4ColSums * colSums(r2X2^2)
check("B1","B4", B1B4s) 
```

$`r term("B1","C1")`$
```{r B1C1}
#B1C1
B1C1s <- sapply(1:nY, function(u) {
  Y4ColSums[u] * xyCors2ColSums[u] * crossprod(r[u,], r3[u,])
}) * (m-1)
check("B1","C1", B1C1s)
```

$`r term("B1","C2")`$
```{r B1C2}
#B1C2
Z <- rX * r2X2
B1C2s <- sapply(1:nY, function(u) {
  Y4ColSums[u] * crossprod(Y[,u], Z[,u])
})
check("B1","C2", B1C2s)
```

$`r term("B1","C3")`$
```{r B1C3}
#B1C3
B1C3s <- B1C1s
check("B1","C3", B1C3s)
```

$`r term("B1","C4")`$
```{r B1C4}
#B1C4
B1C4s <- B1C2s
check("B1","C4", B1C4s)
```

$`r term("B2","B2")`$
```{r B2B2}
#B2^2
B2B2s <- sapply(1:nY, function(u){
  crossprod(r_squared[u,], r2[u,]^2) * sum(xyCors2[,u]) 
}) * (m-1)^2

check("B2","B2", B2B2s)
```

$`r term("B2","B3")`$
```{r B2B3}
#B2B3
B2B3s <- sapply(1:nY, function (u) {
  crossprod(r_squared[u,], r2[u,])^2
}) * (m-1)^2

check("B2","B3", B2B3s)
```

$`r term("B2","B4")`$
```{r B2B4}
#B2B4
rsqr2 <- r_squared * r2
rsqr2X2 <- tcrossprod(X2, rsqr2)
B2B4s <- colSums(rsqr2X2 * r2X2) * (m-1)
check("B2","B4", B2B4s)
```

$`r term("B2","C1")`$
```{r B2C1}
#B2C1
B2C1s <- sapply(1:nY, function(u) {
  crossprod(r_squared[u,], r2[u,]) * crossprod(r[u,],r3[u,])
}) * (m-1)^2
check("B2","C1", B2C1s)
```

$`r term("B2","C2")`$
```{r B2C2}
#B2C2
Z <- rsqr2X2 * rX
B2C2s <- sapply(1:nY, function(u) {
  crossprod(Y[,u], Z[,u])
}) * (m-1)
check("B2","C2", B2C2s)
```

$`r term("B2","C3")`$
```{r B2C3}
#B2C3
B2C3s <- rowSums(r * r2 * r3) * rowSums(r_squared) * (m-1)^2
check("B2","C3", B2C3s)
```

$`r term("B2","C4")`$
```{r B2C4}
#B2C4
Z <- tcrossprod(X, r * r2) * r2X2
B2C4s <- sapply(1:nY, function(u) {
  crossprod(Y[,u], Z[,u])
}) * (m-1)
check("B2","C4", B2C4s)
```
$`r term("B3","B3")`$
```{r B3B3}
#B3^2
B3B3s <- B2B2s
check("B3","B3", B3B3s)
```

$`r term("B3","B4")`$
```{r B3B4}
#B3B4
B3B4s <- B2B4s
check("B3","B4", B3B4s)
```

$`r term("B3","C1")`$
```{r B3C1}
#B3C1
B3C1s <- B2C3s
check("B3","C1", B3C1s)
```

$`r term("B3","C2")`$
```{r B3C2}
#B3C2
B3C2s <- B2C4s
check("B3","C2",B3C2s)
```

$`r term("B3","C3")`$
```{r B3C3}
#B3C3
B3C3s <- B2C1s
check("B3","C3", B3C3s)
```

$`r term("B3","C4")`$
```{r B3C4}
#B3C4
B3C4s <- B2C2s
check("B3", "C4", B3C4s)
```

$`r term("B4","B4")`$
```{r B4B4}
#B4^2
B4B4s <- sapply(1:nY, function(u){
  sum((X2 %*% (r_squared[u,] * tX2))^2)
})
check("B4", "B4", B4B4s)
```

$`r term("B4","C1")`$
```{r B4C1}
#B4C1
B4C1s <- colSums(rr3X2 * r2X2) * (m-1)
check("B4", "C1", B4C1s)
```

$`r term("B4","C2")`$
```{r B4C2}
#B4C2
B4C2s <- sapply(1:nY, function(u){
 crossprod(colSums((X2 %*% (r_squared[u,] * tX2)) * (X2 %*% (r[u,] * tX))),Y[,u])
})
check("B4", "C2", B4C2s)
```

$`r term("B4","C3")`$
```{r B4C3}
#B4C3
B4C3s <- B4C1s
check("B4", "C3", B4C3s)
```


$`r term("B4","C4")`$
```{r B4C4}
#B4C4
B4C4s <- B4C2s
check("B4", "C4", B4C4s)
```

$`r term("C1","C1")`$
```{r C1C1}
#C1^2
C1C1s <- rowSums(r_squared) * rowSums(r3^2) * (m-1)^2
check("C1", "C1", C1C1s)
```

$`r term("C1","C2")`$
```{r C1C2}
#C1C2
Z <- r2X2 * r3X 
C1C2s <- sapply(1:nY, function(u){
  crossprod(Y[,u], Z[,u])
}) * (m-1)
check("C1", "C2", C1C2s)
```

$`r term("C1","C3")`$
```{r C1C3}
#C1C3
C1C3s <- rowSums(r*r3)^2 * (m-1)^2
check("C1", "C3", C1C3s)
```

$`r term("C1","C4")`$
```{r C1C4}
#C1C4
Z <- rX * rr3X2
C1C4s <- sapply(1:nY, function(u){
  crossprod(Y[,u], Z[,u])
}) * (m-1)
check("C1", "C4", C1C4s)
```

$`r term("C2","C2")`$
```{r C2C2}
#C2^2
C2C2s <- sapply(1:nY, function(u){
  crossprod(Y[,u], (XXt * (X2 %*% (r_squared[u,]*tX2))) %*% Y[,u])
})
check("C2", "C2", C2C2s)
```

$`r term("C2","C3")`$
```{r C2C3}
#C2C3
C2C3s <- C1C4s
check("C2", "C3", C2C3s)
```


$`r term("C2","C4")`$
```{r C2C4}
#C2C4
C2C4s <- sapply(1:nY, function(u){
  Z <- X %*% (r[u,] * tX2)
  crossprod(Y[,u], (Z * t(Z)) %*% Y[,u])
})
check("C2", "C4", C2C4s)
```

$`r term("C3","C3")`$
```{r C3C3}
#C3^2
C3C3s <- C1C1s
check("C3", "C3", C3C3s)
```



$`r term("C3","C4")`$
```{r C3C4}
#C3^2
C3C4s <- C1C2s
check("C3", "C4", C3C4s)
```

$`r term("C4","C4")`$
```{r C4C4}
#C4^2
C4C4s <- C2C2s
check("C4", "C4", C4C4s)
```